\Name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Method for a Precision Spectrograph

### 2020-08-25 paper/paper.tex
Debra Comments and RV edits

% typesetting words
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\name}{\project{Excalibur}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\expres}{\project{\acronym{EXPRES}}}
\newcommand{\espresso}{\project{\acronym{ESPRESSO}}}
\newcommand{\neid}{\project{\acronym{NEID}}}
\newcommand{\eprv}{\acronym{EPRV}}
\newcommand{\lfc}{\acronym{LFC}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\target}{HD 34411}

@@

\begin{document}
\title{\name:
  A Non-Parametric, Hierarchical \\
  Wavelength-Calibration Method for a Precision Spectrograph}

@@

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
\affil{Flatiron Institute, Simons Foundation, 162 Fifth Avenue, New York, NY 10010, USA}
\affil{Center for Data Science, New York University, 60 Fifth Avenue, New York, NY 10011, USA}
\affil{Max-Planck-Institut für Astronomie, Königstuhl 17, D-69117 Heidelberg, Germany}

@@

\begin{abstract}\noindent%
\name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  This method is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ of a pixel.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments with highly stabilized optical systems and detectors.  \name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all possible calibration states for an instrument.  \name\ also takes advantage of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities, more than any parametric model.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which includes a laser frequency comb.  \name\ wavelengths predicts the wavelengths of known calibration lines at smaller per-line RMS by a factor of $\sim$5 when compared to a classic, polynomial wavelength calibration that treats each exposure individually.  For ten months of radial-velocity measurements of HD 34411, using \name\ wavelengths reduced the RMS of the complete data set from $1.22\, \mps$ to $1.05\, \mps$; this improvement is equivalent to reducing the wavelength term in the error budget by $0.62\,\mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}

\section{Introduction} 
Precise, radial-velocity (\eprv) programs have been incredibly fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $10^5$, which correspond to line widths on the order of $3000\,\mps$.  Exoplanet science happens at the level of $3\,\mps$ precision, which is sensitive enough to discover gas-giant planets around other stars.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).    Some hardware--software systems, such as \espresso, \expres,  \neid, etc., were designed to deliver precision at this level \citep{pepe2013,  jurgenson2016, neid}.  Early data from \expres\ and \espresso\ suggest that the hardware improvements truly propagate through to sub-$\mps$ uncertainties on final, on-sky RV measurements \citep{blackman2020, petersburg2020, mascareno2020}.

@@

Here, we propose to simplify and improve calibration programs for \eprv\ hardware systems with two practical yet innovative ideas.  The first flows from the fact that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.

This recommends a calibration methodology that is \emph{non-parametric}, or not defined by a prescribed, analytic function described by a finite number of parameters :  If every point in the spectrograph detector is sufficiently surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous ninth-order polynomial, for example).  In some ways, this is a generalization of the recent work that demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  Going non-parametric will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials are at detector edges).

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  This is particularly likely with the latest generation of \eprv\ instruments that are equipped with stringent environmental stabilizing, likely reducing the number and extent of the axes along which the instrument can vary.  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the small, accessible part of calibration space the spectrograph was located for each exposure.

@@

\name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, primarily because the need for good wavelength calibration is so great in this field.  Irregardless, we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph, though the precision of the returned wavelengths will depend on the available calibration sources (more discussion in Section \ref{sec:others} below) .

\section{Method} \label{sec:method}
The \name\ method is designed to take a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas; The wavelength solution should be given enormous flexibility, but it lives in a very low-dimensional calibration space, where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional while assuming very little about the forms of those states.

\name\ also assumes dense enough calibration line coverage with well-fit line centers to provide sufficient constraint on an interpolated wavelength solution across an echelle order.  Upstream errors in line center positions may propagate through \name\ wavelength models.  The needed line density is dependent on the required precision of the returned wavelength model; larger spacing between lines offer less constraint and are likely to return worse wavelengths.  We revisit and quantify these conditions in Section \ref{sec:others}.

Wavelength calibration is usually posed in the following way.  Given an exposure $n$, and echelle order $m$, there is a relationship between
the two-dimensional $(x,y)$-position on the detector and the
wavelength $\lambda$
\begin{equation}
\lambda(x,y,m,n) = f(x,y,m;\theta_{n})
\quad ,
\label{eq:wsol}
\end{equation}
where $\theta_{n}$ represents the parameters describing the wavelength solution for a given exposure.

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here, the y-dependence is dropped as that dependence is carried by spectral order $m$.  The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated from the time of calibration exposures to time $t_n$ of a science exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration, science exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

Given a stabilized instrument with low degrees of freedom, however,  the calibration of any image can and should be informed by the calibration of every other image.  The calibration data themselves can be used to develop a low-dimensional basis for expressing the space of all possible calibrations for a spectrograph with few degrees of freedom.

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  The resultant $x(\lambda,m,n)$, a list of calibration line positions for a given exposure, can be regarded as the calibration state of the spectrograph for that exposure.  When this calibration structure is used to deliver a wavelength solution, $x(\lambda,m,n)$ will be inverted back into $\lambda(x,m,n)$ to recover wavelengths for each detector position $x$ and echelle order $m$..

The challenge is to learn these basis functions from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern these basis functions.  In this paper, we present a model using principal component analysis (PCA) \citep{wiki_pca}.  A PCA is justifiable \lz{(Hogg)} in the limit where there is very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{A cartoon representation of the \name\ method, as described in Section \ref{sec:method}.  We exaggerate variations in measured line position, changes in calibration space, and interpolation deviations for clarity.  In step one, dimensionality reduction and denoising (\textsection \ref{sec:denoising}), the complete set of line positions for all exposures is analyzed to return a set of $K$ basis vectors, $g_n$, which represent different ways the spectrograph calibration changes.  These basis vectors span the $K$-dimensional calibration space of the spectrograph, which includes all possible wavelength solutions.  In step two (\textsection \ref{sec:interp_time}), \name\ interpolates the amplitudes of each basis vector, $a_n,k$, to return the calibration state for a specific science exposure, returned as a set of de-noised calibration lines.  The assigned wavelengths of these de-noised line positions are then interpolated onto other pixels in step three (\textsection \ref{sec:interp_wsol}) to construct a full wavelength model that returns wavelength as a function of detector position $x$ and echelle order $m$.}
\label{fig:cartoon}
\end{figure*}

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
\name\ will use calibration images to  1) determine the complete calibration space in which an instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.  There are many strategies for identifying calibration line positions and matching them to their assigned wavelengths; this problem is left out of scope for this work.  However, we caution that systematic errors or large uncertainties in fitting line positions can propagate through to the wavelength models returned by \name.

For each exposure, $n$, every line, $(\lambda,m)$, has an associated fitted detector position, $x(\lambda,m,n)$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a \code{NaN} (hardware not-a-number) for that echelle order instead.  Let there be $P$ lines per exposure.  \name\ reads in a $N \times P$ matrix of line positions for each of $P$ lines for each of $N$ exposures.

The mean of measured line position over the set of calibration exposures represents the fiducial, or standard calibration of the spectrograph, $g_0(\lambda,m)$.  In this implementation of \name, principal component analysis is performed on the difference between this fiducial calibration and each individual line position.  The returned principal components serve as basis functions,  $g_k(\lambda,m)$, expressing the possible deviations of the spectrograph from this fiducial calibration.  The magnitude of each principal component for each exposure, $a_{nk}$, represents the scalar amplitude of these deviations for each exposure.  \name\ then uses a small number, $K$, of principal components to reconstruct a de-noised version of the line positions as formulated in Equation \ref{eq:excl_wsol}.

@@

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Basis vectors of the low-dimensional calibration space $g_k(\lambda,m)$, location of exposures in calibration space $a_{n,k}$}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ for $x(\lambda,m,n)$ = \code{NaN} where $K$ is a a small integer
	}
\caption{Dimension Reduction and De-Noising}
\end{algorithm}

\subsection{Interpolating Calibration Position}
 \label{sec:interp_time}
\name\ then interpolates the amplitude, $a_{n,k}$, of each principal component to determine the calibration state of the spectrograph.  For example, the amplitude can be interpolated with respect to time to recreate the calibration state of the spectrograph at different times.  The choice of what to interpolate against depends on the dominant contribution to variation in the calibration of the instrument.

In the implementation of \name\ presented here, the amplitudes of the principal components are interpolated linearly with respect to time.  \name\ finds new magnitudes $a_{n'k}$ for a science exposure $n'$ at time $t_{n'}$, which are used to construct the calibration state of the spectrograph for that point in time.  Using interpolated magnitudes, $a_{n,k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.

\subsection{Interpolating a Wavelength Solution} \label{sec:interp_wsol}
From the de-noising step, \name\ can now construct a set of calibration lines, $x(\lambda,m,n')$ for any exposure $n'$.  To construct a wavelength solution, \name\ interpolates the known wavelengths over the exposure specific line centers.  For instance, interpolating the known wavelengths vs. line centers onto every integer x will generate wavelengths for each pixel in an echelle order.

We found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a continuous function, while the enforced monotonicity prevents large deviations that may befall a cubic spline.  Choice in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

\section{Data} \label{sec:data}
We tested \name\ using data from \expres, the EXtreme PRecison Spectrograph.  \expres\ is an environmentally-stabilized, fiber-fed, $R\sim137,000$, optical spectrograph \citep{jurgenson2016, blackman2020}.  \expres\ has two different wavelength calibration sources, a ThAr lamp and a Menlo Systems laser frequency comb (LFC).  LFCs are unique in that the wavelengths of their emission lines are stable and exactly known to the order of pico-meters  \citep{wilken2012, molaro2013, probst2014}.

Rather than using a simultaneous calibration fiber, two to three LFC exposures are interspersed between science exposures roughly every 30 minutes while the telescope is slewing to different targets.  ThAr exposures are taken at the beginning and end of each night.  All calibration data is taken through the science fiber, so that calibration light travels down the same optical pathway and is projected onto the same pixels as all science observations.

LFC lines cover echelle orders 84-124, which contain 19203 calibration lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.   ThAr lines cover all 86 extracted orders of \expres\ (echelle orders 75-160), which include 5295 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.

 \name\ inputs a list of echelle orders $m$, line wavelengths $\lambda$, and pixel positions $x$ of each line for each exposure.  We, therefore, use line positions generated by the pre-existing \expres\ pipeline from the following procedure \citep{petersburg2020}.

>>>>

% typesetting words
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\name}{\project{excalibur}}
\newcommand{\Name}{\project{Excalibur}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\expres}{\project{\acronym{EXPRES}}}
\newcommand{\espresso}{\project{\acronym{ESPRESSO}}}
\newcommand{\neid}{\project{\acronym{NEID}}}
\newcommand{\eprv}{\acronym{EPRV}}
\newcommand{\lfc}{\acronym{LFC}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\target}{HD 34411}

@@

\begin{document}
\title{\Name:
  A Non-Parametric, Hierarchical \\
  Wavelength-Calibration Method for a Precision Spectrograph}

@@

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
\affil{Flatiron Institute, Simons Foundation, 162 Fifth Avenue, New York, NY 10010, USA}
\affil{Center for Data Science, New York University, 60 Fifth Avenue, New York, NY 10011, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, Königstuhl 17, D-69117 Heidelberg, Germany}

@@

\begin{abstract}\noindent%
\Name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  It is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ pixels.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments, which feature highly stabilized optical systems and detectors.  \Name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all possible calibration states for an instrument.  \Name\ also takes advantage of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities, better than any parametric model, such as a polynomial.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which includes a laser frequency comb.  When wavelengths are assigned to laser comb lines using \name, the RMS of the residuals is about five times lower than wavelengths assigned using polynomial fits to individual exposures.  In a typical test case, radial-velocity measurements of HD~34411 showed a reduction in RMS scatter over a 10-month time baseline from $1.22$ to $1.05\, \mps$.  This improvement is equivalent to removing an independent, additive error of $0.62\, \mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}
\section*{~}\pagebreak
\section{Introduction} 
Precise, radial-velocity (\eprv) programs have been fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $50,000-100,000$, which correspond to line widths on the order of $3000\,\mps$.  The state of the art RV precision reached $1\, \mps$ in 2016.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).  Both \expres\ and \espresso\ have been commissioned for more than a year and are demonstrating $<0.1 \mps$ instrumental errors and $\sim 0.2\, \mps$ errors on stars \citep{pepe2013, jurgenson2016, blackman2020, peterburg2020, brewer2020, mascareno2020}.

@@

Here, we propose to simplify and improve calibration programs for \eprv\ hardware systems with two practical yet innovative ideas.  The first flows from the fact that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.  This enables use of a calibration methodology that is \emph{non-parametric}, or not defined by a prescribed, analytic function described by a finite number of parameters :  If every point in the spectrograph detector is sufficiently surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous ninth-order polynomial, for example).  In some ways, this is a generalization of recent work, which has demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  A non-parametric approach will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials are at detector edges).

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  This is particularly true of \eprv\ instruments, which are equipped with stringent environmental stabilizing, likely reducing the number and extent of the axes along which the instrument can vary.  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the small, accessible part of calibration space the spectrograph was located for each exposure.

@@

The method \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, primarily because the need for good wavelength calibration is so great in this field.  Irregardless, we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \Name\ should be applicable to other spectrograph with low-dimensional variability, though the precision of the returned wavelengths will depend on the available calibration sources (more discussion in Section \ref{sec:others} below) .

\section{Method} \label{sec:method}
The \name\ method is designed to take many calibration images, each containing a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  It operates on the two core ideas; The wavelength solution should be allowed flexibility, but it lives in a very low-dimensional calibration space, where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \Name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional while assuming very little about the forms of those states.

\Name\ assumes dense-enough calibration line coverage with well-fit line centers to provide sufficient constraints on an interpolated wavelength solution across an echelle order.  Upstream errors in line center positions may propagate through \name\ wavelength models.  The required line density is dependent on the required precision of the returned wavelength model; larger spacing between lines offer less constraint and are likely to return worse wavelengths.  We revisit and quantify these conditions in Section \ref{sec:others}.

Wavelength calibration is usually posed in the following way: Given an exposure $n$, and echelle order $m$, there is a relationship between
the two-dimensional $(x,y)$-position on the detector and the
wavelength $\lambda$
\begin{equation}
\lambda(x,y,m,n) = f(x,y,m;\theta_{n})
\quad ,
\label{eq:wsol}
\end{equation}
where $\theta_{n}$ represents the parameters describing the wavelength solution for a given exposure.

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to a 2D, 9\textsuperscript{th}-order polynomial, where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here, the y-dependence is dropped as, in our application, this dependence is carried by spectral order $m$.  The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated from the time of calibration exposures to time $t_n$ of a science exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration, science exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

Given a stabilized instrument with low degrees of freedom, however, the calibration of any image can be reliably informed by the calibration of every other image.  The calibration data themselves can be used to develop a low-dimensional basis for expressing the space of all possible calibrations for a spectrograph with few degrees of freedom.

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  The resultant $x(\lambda,m,n)$, a list of calibration line positions for a given exposure, can be regarded as the calibration state of the spectrograph for that exposure.  When this calibration structure is used to deliver a wavelength solution, $x(\lambda,m,n)$ can be inverted back into $\lambda(x,m,n)$ to recover wavelengths for each detector position $x$ and echelle order $m$..

The challenge is to learn these basis functions, $g_k(\lambda,m)$, from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern the basis functions.  In this paper, we present a model using principal component analysis (PCA) \citep{wiki_pca}.  A PCA is justifiable in the limit where there is very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{A cartoon representation of the \name\ method, as described in Section \ref{sec:method}.  We exaggerate variations in measured line position, changes in calibration space, and interpolation deviations for clarity.  In step one, dimensionality reduction and denoising (\textsection \ref{sec:denoising}), the complete set of line positions for all exposures is analyzed to return a set of $K$ basis vectors, $g_n$, which represent different ways the spectrograph calibration changes.  These basis vectors span the $K$-dimensional calibration space of the spectrograph, which includes all possible wavelength solutions.  In step two (\textsection \ref{sec:interp_time}), the amplitude of each basis vector, $a_{n,k}$, is interpolated to return the calibration state for a specific science exposure, returned as a set of de-noised calibration lines.  The assigned wavelengths of these de-noised line positions are then interpolated onto other pixels in step three (\textsection \ref{sec:interp_wsol}) to construct a full wavelength model that returns wavelength as a function of detector position $x$ and echelle order $m$.}
\label{fig:cartoon}
\end{figure*}

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
\Name\ will use calibration images to determine 1) the space in which an instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.  There are many strategies for identifying calibration line positions and matching them to their assigned wavelengths; this problem is left out of scope for this work.  That is, \name\ assumes that this has been done correctly.  We caution, though, that systematic errors or large uncertainties in fitting line positions can propagate through to errors or biases in the wavelength models returned by \name.

For each exposure, $n$, every line, $(\lambda,m)$, has an associated fitted (measured) detector position, $x(\lambda,m,n)$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a \code{NaN} (hardware not-a-number) for that echelle order instead.  Let there be $P$ lines per exposure.  \Name\ reads in a $N \times P$ matrix of line positions for each of $P$ lines for each of $N$ exposures.

The mean of measured line position over the set of calibration exposures represents the fiducial, or standard calibration of the spectrograph, $g_0(\lambda,m)$.  In this implementation of \name, principal component analysis is performed on the difference between this fiducial calibration and each individual line position.  The returned principal components serve as basis functions,  $g_k(\lambda,m)$, expressing the possible deviations of the spectrograph from this fiducial calibration.  The magnitude of each principal component for each exposure, $a_{nk}$, represents the scalar amplitude of these deviations for each exposure.  \Name\ then uses a small number, $K$, of principal components to reconstruct a de-noised version of the line positions as formulated in Equation \ref{eq:excl_wsol}.

@@

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Basis vectors of the low-dimensional calibration space $g_k(\lambda,m)$, location of exposures in calibration space expressed by amplitudes $a_{n,k}$}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ for $x(\lambda,m,n)$ = \code{NaN} where $K$ is a a small integer
	}
\caption{Dimensionality Reduction and De-Noising}
\end{algorithm}

\subsection{Interpolating Calibration Position}
 \label{sec:interp_time}
In \name\, the amplitude, $a_{n,k}$, of each principal component is then interpolated to determine the calibration state of the spectrograph.  For example, the amplitude can be interpolated with respect to time to recreate the calibration state of the spectrograph at different times.  The choice of what to interpolate against depends on the dominant contribution to variation in the calibration of the instrument.

In the implementation of \name\ presented here, the amplitudes of the principal components are interpolated linearly with respect to time.  These new, linearly interpolated magnitudes, $a_{n\,k}$ for a science exposure $n'$ at time $t_{n'}$ are then used to construct the calibration state of the spectrograph for that point in time.  Using interpolated magnitudes, $a_{n,k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, a new set of calibration lines, $x(\lambda,m,n')$ can be constructed
 for any exposure as prescribed in Equation \ref{eq:excl_wsol}.

\subsection{Interpolating a Wavelength Solution} \label{sec:interp_wsol}
From the de-noising step, \name\ can now construct a set of calibration lines, $x(\lambda,m,n')$ for any exposure $n'$.  To construct a wavelength solution, we invert $x(\lambda,m,n')$ to $\lambda(x,m,n')$ by interpolating known wavelengths of the calibration lines over detector position.  For instance, interpolating the known wavelengths vs. line centers onto every integer x will generate wavelengths for each pixel in an echelle order.

We found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a continuous function, while the enforced monotonicity allows the wavelength solution, $\lambda(x,m,n')$, to be invertible and prevents large deviations that may befall a cubic spline.  Choice in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

\section{Data} \label{sec:data}
We tested \name\ using data from \expres, the EXtreme PRecison Spectrograph.  \expres\ is an environmentally-stabilized, fiber-fed optical spectrograph with a median resolving power $R=\lambda/\delta_{\lambda}=\sim137,000$, over the wavelength range of $390-780\, nm$ \citep{jurgenson2016, blackman2020}.  \expres\ has two different wavelength calibration sources, a ThAr lamp and a Menlo Systems laser frequency comb (LFC).  LFCs are unique in that the wavelengths of their emission lines are stable and exactly known to the order of pico-meters  \citep{wilken2012, molaro2013, probst2014}.

Rather than using a simultaneous calibration fiber, two to three LFC exposures are obtained roughly every 30 minutes while the telescope is slewing to the next target.  ThAr exposures are taken at the beginning and end of each night.  All calibration data is taken through the science fiber, so that calibration light travels down the same optical pathway and is projected onto the same pixels as all science observations.

LFC lines cover echelle orders 84-124, which contain approximately 19200 calibration lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.   ThAr lines cover all 86 extracted orders of \expres\ (echelle orders 75-160), which include approximately 5300 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.  A list of echelle orders $m$, line wavelengths $\lambda$, and pixel positions $x$, which have been calculated by the EXPRES pipeline \citep{petersburg2020}, are read into \name.
